---
output: github_document
bibliography: "./vignettes/articles/Wright Replication.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GWFdata

<!-- badges: start -->
<!-- badges: end -->

GWFdata archives the data used in @wrightLatentCharacteristicsThat2021 and @geddesHowDictatorshipsWork2018. It also includes a subset of the personalism indicators in @weeksDictatorsWarPeace2014. The primary purpose of the package is to replicate the IRT model described in @wrightLatentCharacteristicsThat2021, based on the Stata .do file archived at the [Political Science Research and Methods dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/3BC21O). An article with detailed instructions on how to do this is [included in this website](https://xmarquez.github.io/GWFdata/articles/replicating-wright-2019.html).  

## Installation

You can install the development version of GWFdata from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("xmarquez/GWFdata")
```

## Examples

This is a data-only package. Just type the name of the dataset you want. The main dataset is `gwf`.

```{r example}
library(GWFdata)
gwf
```

Documentation on the included variables be accessed by typing `?gwf`.

The package also includes the dataset `weeks` and `gwf_irt`. The latter includes the replicated factor scores and IRT estimates reported in @wrightLatentCharacteristicsThat2021.

## References

